---
format: "html"
bibliography: "references.yaml"
---

# Resolving taxon names of Japanese ferns: A case study

## Introduction

This analysis shows how the [taxastand](https://github.com/joelnitta/taxastand) R package can be used to join datasets by resolving species names to a common, custom taxonomy.

The goal is to generate a heatmap of endangered fern species in Japan.

## Methods

There are two data sources:

- [GBIF](https://gbif.org), which includes distribution data (occurrence records) 

- [Green List](https://doi.org/10.5061/dryad.4362p32), which includes conservation status for the ferns of Japan.

However, these data sources use different taxonomies, so names may not match between them. So, name harmonization is done by resolving names in each of the two sources to the [pteridocat](https://github.com/fernphy/pteridocat) taxonomic database first, then merging.

```{r}
#| label: load-data
#| echo: false
#| message: false

source("packages.R")

tar_load(c(
  ja_ferns_gbif_accepted,
  gbif_names_resolved_pt,
  gbif_names_resolved_mapped,
  mesh_species,
  japan_mesh2,
  mesh_rich,
  green_names_resolved_pt_single,
  green_list_no_hybrids,
  ja_gbif_names,
  gbif_names_resolved_pt_single))
```

## Results

### Name resolution

```{r}
#| label: match-stats
#| echo: false
#| results: "hide"

# Counts
green_list_species_n <- green_list_no_hybrids %>%
  pull(sci_name) %>%
  n_distinct()

green_list_species_resolved_n <- green_names_resolved_pt_single %>%
  pull(query) %>%
  n_distinct()

gbif_species_n <- ja_gbif_names %>%
  pull(scientificName) %>%
  n_distinct()

gbif_species_resolved_n <- gbif_names_resolved_pt_single %>%
  pull(query) %>%
  n_distinct()
```

Of `r green_list_species_n` scientific names in the Green List (hybrid formulas excluded), `r green_list_species_resolved_n` names were successfully resolved.

Of `r gbif_species_n` scientific names in the GBIF data, `r gbif_species_resolved_n` names were successfully resolved.

`r nrow(gbif_names_resolved_mapped)` species were successfully mapped from GBIF data to Green List data via pteridocat, including `r nrow(mesh_species)` occurrences (@tbl-gbif-summary).

```{r}
#| label: tbl-gbif-summary
#| tbl-cap: "Types of matches for species names in GBIF successfully resolved to pteridocat"
#| tbl-colwidths: [80,20]
#| echo: false
#| message: false

table <-
gbif_names_resolved_mapped %>%
  filter(!is.na(ja_name)) %>%
  select(query = gbif_name) %>%
  unique() %>%
  left_join(
    unique(select(gbif_names_resolved_pt, query, match_type)),
    by = "query") %>%
  assert(is_uniq, query) %>%
  count(match_type) %>%
  mutate(
    match_group = case_when(
      match_type %in% c(
        "auto_basexin", "auto_cfonly",
        "auto_exin", "auto_exin-", "auto_exin+") ~ "Taxonomic rule",
      str_detect(match_type, "auto_noauth") ~ "Missing author",
      str_detect(match_type, "punc") ~ "Difference in punctuation",
      str_detect(match_type, "fuzzy") ~ "Fuzzy",
      str_detect(match_type, "exact") ~ "Exact match",
    )
  ) %>%
  group_by(match_group) %>%
  summarize(
    n = sum(n)
  ) %>%
  arrange(desc(n))

kable(table)
```

Plots of species richness (@fig-richness) and endangered species richness (@fig-endangered) appeared mostly similar to a previous study using data not obtained from GBIF [@Nitta2022a], supporting the reliability of this method. However, richness of endangered species (@fig-endangered) appeared slightly higher than in @Nitta2022a, which could be due to artifacts in GBIF data (false positives). Since @fig-endangered is shown on a log-scale, even a very few number of false positives will clearly appear on the map.

```{r}
#| label: fig-richness
#| echo: false
#| message: false
#| warning: false
#| fig-cap: Species richness of all Japanese ferns, GBIF data matched to Green List via pteridocat

ferns_rich_plot <-
  mesh_rich %>%
  mutate(rich = replace_na(rich, 0)) %>%
  ggplot() +
  geom_sf(
    aes(fill = rich),
    color = "transparent") +
  scale_fill_scico(
    "No. species",
    palette = "lajolla",
    na.value = "grey80",
    labels = function(x) scales::number(x, accuracy = 1),
    breaks = c(0, 50, 100, 150, 200, 250)) +
  scale_x_continuous(breaks = c(130, 140), limits = c(122,147.4)) +
  ylim(c(24, 46)) +
  theme_gray(base_size = 16) +
  theme(
    panel.grid.major = element_line(size = 0.1, color = "dark grey"),
    panel.grid.minor = element_blank(),
    plot.subtitle = element_text(face = "bold"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    legend.title = element_text(size = 20 * 1.5 / .pt),
    legend.text = element_text(size = 16 * 1.5 / .pt),
    legend.justification = c(0, 1), legend.position = c(0, 1)
  )

ggsave(
  plot = ferns_rich_plot, file = "results/ferns_rich.png",
  height = 5, width = 5, units = "in")

ferns_rich_plot
```

```{r}
#| label: fig-endangered
#| echo: false
#| message: false
#| warning: false
#| error: false
#| fig-cap: Species richness of endangered Japanese ferns, GBIF data matched to Green List via pteridocat

endangered_ferns_rich_plot <-
  mesh_rich %>%
  mutate(rich_endan = replace_na(rich_endan, 0)) %>%
  ggplot() +
  geom_sf(
    aes(fill = rich_endan),
    color = "transparent") +
  scale_fill_scico(
    "No. species\n(log scale)",
    trans = "log",
    palette = "lajolla", na.value = "grey80",
    oob = scales::squish_infinite,
    labels = function(x) scales::number(x, accuracy = 1),
    breaks = c(3, 6, 12, 24, 48)) +
  scale_x_continuous(breaks = c(130, 140), limits = c(122,147.4)) +
  ylim(c(24, 46)) +
  theme_gray(base_size = 16) +
  theme(
    panel.grid.major = element_line(size = 0.1, color = "dark grey"),
    panel.grid.minor = element_blank(),
    plot.subtitle = element_text(face = "bold"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    legend.title = element_text(size = 20 * 1.5 / .pt),
    legend.text = element_text(size = 16 * 1.5 / .pt),
    legend.justification = c(0, 1), legend.position = c(0, 1)
  )

ggsave(
  plot = endangered_ferns_rich_plot, file = "results/endangered_ferns_rich.png",
  height = 5, width = 5, units = "in")

suppressWarnings(
  endangered_ferns_rich_plot)
```